

# Spatial spot-deconvolution with SCDC


## Load libraries

```{r}

library(Seurat)
library(SCDC)
library(SingleCellExperiment)
library(biomaRt)
library(Matrix)
library(niceRplots)
library(rhdf5)
library(rafalib)
library(harmony)
library(future)
library(future.apply)
library(dplyr)
library(WGCNA)


remotes::install_github('czarnewski/niceRplots')

```


## Compute the top DGE from single-cell reference

```{r}

sc_seurat <- readRDS("GSE148794_single_cell_seurat.rds")
sample_size <- table(sc_seurat$clusters_louvain)

sample_size[ sample_size > 20 ] <- 20
sample_size

DGE_cells <- lapply(names(sample_size), function(x){ 
  set.seed(1)
  sample( colnames(sc_seurat) [ sc_seurat$clusters_louvain == x ] , size = sample_size[x])
  })
DGE_cells <- unlist(DGE_cells)


DGE_DATA <- sc_seurat[, DGE_cells]
DGE_DATA <- SetIdent( DGE_DATA , value = "clusters_louvain")
```



## Compute SCDC deconvolution 

```{r}
sc_eset_SC <- ExpressionSet(assayData=as.matrix(DGE_DATA@assays$RNA@counts[m_feats,]),
 phenoData =  AnnotatedDataFrame(DGE_DATA@meta.data))
CRC_eset <- ExpressionSet(assayData=as.matrix(DATA@assays$RNA@counts[m_feats,]),
 phenoData = AnnotatedDataFrame(DATA@meta.data))


deconvolution_crc <- SCDC::SCDC_prop(bulk.eset = CRC_eset,
 sc.eset = sc_eset_SC,
 ct.varname = "clusters_louvain",
 ct.sub = as.character(unique(sc_eset_SC$clusters_louvain)) )
saveRDS(deconvolution_crc , "deconvolution_crc.rds" )

DATA@assays[["SCDC_prop"]] <- Seurat::CreateAssayObject(data = t(deconvolution_crc$prop.est.mvw),
 min.cells = 0,min.features = 0)
DATA@assays[["SCDC_prop_Cluster"]] <- Seurat::CreateAssayObject(data = rowsum(DATA@assays$SCDC_prop@data, 
 gsub("_Healthy|_Inflamed|_Non[-]inflamed","",rownames(DATA@assays$SCDC_prop@data))),
 min.cells = 0,min.features = 0)
saveRDS(DATA,"Visium_dataset_CRC.rds")
```



## Load single-cell annotation onto the spatial

```{r}
DATA <- readRDS("Visium_dataset_CRC.rds")
sc_seurat <- readRDS("GSE148794_single_cell_seurat.rds")

annot <- read.csv2("CellTypeAnnot.csv",row.names = 1)

sc_seurat$annot <- annot$fine_annotation[match(sc_seurat$clusters_louvain,rownames(annot))]

tmp <- rowsum( DATA@assays$SCDC_prop@data , annot$fine_annotation )
DATA@assays[["SCDC_prop_annot"]] <- Seurat::CreateAssayObject(data = tmp,  min.cells = 0,min.features = 0)

img_list <- names(DATA@images)
clusters <- rownames(DATA@assays$SCDC_prop_annot@data)

```


## Plot the cell type labels from annotation

```{r}

if(!dir.exists("SCDC_annot")){dir.create("SCDC_annot",recursive = T)}

pdf("SCDC_annot/_HE.pdf",width = (length(img_list) + 1)*3.5,height = 3.5,useDingbats = F)
mypar(1,length(img_list) + 1)
plot_meta( sc_seurat , feat = "annot" , cex=.5,label = T )
for(i in img_list){
  plot_spatial_feat(DATA, red = i,feat =  "Actb", main="",assay="RNA",
    pch=16,cex=1,plot_tissue = T,
    transparency = "00")}
dev.off()


for(j in clusters){
  pdf(paste0("SCDC_annot/cluster",j,".pdf"),
      width = (length(img_list) + 1)*3.5,
      height = 3.5,useDingbats = F)
  
  mypar(1,length(img_list) + 1)
  sc_seurat$temp <- (sc_seurat$annot == j)*1
  plot_feat( sc_seurat , feat = "temp" , cex=.5, main=j )
  
  for(i in img_list){
    plot_spatial_feat(DATA, 
       red = i, 
        feat =  j, 
        main=paste0(i," (",j,")"),
        assay="SCDC_prop_annot",
        pch=16,cex=1, 
        plot_tissue = F)
  }
dev.off()
}

```



## Compute cell-cell co-detection rate

This is done as from  Frede, Czarnewski et al (2022) Immunity
https://doi.org/10.1016/j.immuni.2022.11.002

```{r}
coocurrence <- lapply(img_list,function(x){
  temp <- DATA@assays$SCDC_prop_annot@data[,DATA$orig.ident == x]
  temp <- sqrt( temp %*% t(temp) / ncol(temp))
  rownames(temp) <- rownames(DATA@assays$SCDC_prop_annot@data)
  colnames(temp) <- rownames(DATA@assays$SCDC_prop_annot@data)
  return( temp )
})
names(coocurrence) <- img_list
```


## Compute cell-cell co-detection rates

```{r}
pdf("cell-cell_codetection_annot.pdf",width = 7*3.5,height = 3*3.5, useDingbats = F)
mypar(3,7,mar=c(1,1,1,1))

plot_meta( sc_seurat,red = "umap",feat = "annot",cex=.8,label = T)
for(i in names(coocurrence)){
  l <-  t(sapply( as.character(unique(sc_seurat$annot)) , 
                          reds=as.data.frame(sc_seurat@reductions$umap@cell.embeddings[,1:2]), 
                          cl1=sc_seurat$annot, function(jj,reds,cl1) { pmean(reds[cl1==jj,])  }))
  plot_meta( sc_seurat,red = "umap",feat = "annot",cex=.8,label = F,col = "grey90",main=i)
  g <- igraph::graph_from_adjacency_matrix( coocurrence[[i]] , mode = "undirected",weighted = T)
  el <- data.frame(as_edgelist(g),E(g)$weight)
  el <- el[el[,3] > 0.01,]
  
  segments( x0 = l[el[,1],1] , y0 = l[el[,1],2] ,x1 =  l[el[,2],1] , y1 = l[el[,2],2],
            col=paste0(colorRampPalette(c("grey80","black"))( 100 ),sprintf("%02d",round((0:99)/2+50) ))[round(el[,3]*98)+1],
            lwd=el[,3]*20)
}


plot_meta( sc_seurat,red = "umap",feat = "annot",cex=.8,label = T)
for(i in names(coocurrence)){
  l <-  t(sapply( as.character(unique(sc_seurat$annot)) , 
                          reds=as.data.frame(sc_seurat@reductions$umap@cell.embeddings[,1:2]), 
                          cl1=sc_seurat$annot, function(jj,reds,cl1) { pmean(reds[cl1==jj,])  }))
  plot_meta( sc_seurat,red = "umap",feat = "annot",cex=.8,label = F,col = "grey90",main=i)
  temp <- log2( ( coocurrence[[i]] +1) /  (coocurrence[[1]] +1) )
  if(sum(abs(temp)) != 0){
      g <- igraph::graph_from_adjacency_matrix( temp , mode = "undirected",weighted = T)
      el <- data.frame(as_edgelist(g),E(g)$weight)

    el <- el[abs(el[,3]) > 0.01,]
    segments( x0 = l[el[,1],1] , y0 = l[el[,1],2] ,x1 =  l[el[,2],1] , y1 = l[el[,2],2],
              col=paste0(colorRampPalette(c("navy","grey80","firebrick"))( 100 ), sprintf("%02d",c(49:0,0:49)+50 ))[ round((el[,3]*200))+50 ] ,
              lwd=abs(el[,3]*20) )
  }
}

plot_meta( sc_seurat,red = "umap",feat = "annot",cex=.8,label = T)
for(i in names(coocurrence)){
  l <-  t(sapply( as.character(unique(sc_seurat$annot)) , 
                          reds=as.data.frame(sc_seurat@reductions$umap@cell.embeddings[,1:2]), 
                          cl1=sc_seurat$annot, function(jj,reds,cl1) { pmean(reds[cl1==jj,])  }))
  plot_meta( sc_seurat,red = "umap",feat = "annot",cex=.8,label = F,col = "grey90",main=i)
  temp <- log2( ( coocurrence[[i]] +1) /  (coocurrence[[3]] +1) )
  if(sum(abs(temp)) != 0){
      g <- igraph::graph_from_adjacency_matrix( temp , mode = "undirected",weighted = T)
      el <- data.frame(as_edgelist(g),E(g)$weight)

    el <- el[abs(el[,3]) > 0.01,]
    segments( x0 = l[el[,1],1] , y0 = l[el[,1],2] ,x1 =  l[el[,2],1] , y1 = l[el[,2],2],
              col=paste0(colorRampPalette(c("navy","grey80","firebrick"))( 100 ), sprintf("%02d",c(49:0,0:49)+50 ))[ round((el[,3]*200))+50 ] ,
              lwd=abs(el[,3]*20) )
  }
}
dev.off()

```


## Differential cell abundance

```{r}
library(glmnet)
library(glmnetUtils)

DATA$orig.ident <- factor(DATA$orig.ident)
temp <- DATA@assays$SCDC_prop_annot@data[,grepl("d22|d43",DATA$orig.ident)]
group <- factor(gsub(".*[_]","",DATA$orig.ident[grepl("d22|d43",DATA$orig.ident)]))
treat <- factor(gsub("d22|d43","chow",DATA$orig.ident[grepl("d22|d43",DATA$orig.ident)]))

plan(multiprocess, workers=future::availableCores()-1  )

temp2 <- data.frame(group,treat)
cl <- future_lapply(rownames(temp),temp=temp,temp2=temp2, function(x,temp,temp2) {
  df <- cbind(temp2,y=temp[x,])
  m <- summary( glm( y ~ group*treat ,data = df) )
  return( m$coefficients )
} , future.seed = 1 )
names(cl) <- rownames(temp)
cl


pdf("differential_abundance_plot_annot.pdf",width = 4*3,height = 4,useDingbats = F)
mypar(1,3)
for( i in rownames(cl[[1]])[-1]){
  
  cl_group <- data.frame(t(sapply(cl,i=i,function(x,i){x[i, c("Estimate","Pr(>|t|)") ]})))
  cl_group <- cbind(cl_group , sig = ( abs(cl_group[,1]) > 0.015 ) & ( cl_group[,2] < 1e-50 ) )
  cl_group[,2][cl_group[,2]==0] <- min(cl_group[,2][cl_group[,2]!=0])

  plot( cl_group[,1] , -log10(cl_group[,2]) , ylim=c(0,400),
        xlim=c(-0.4,0.4), pch=16, main=i, 
        ylab="-log10(pvalue)",xlab="Estimate",
        col= ifelse(cl_group[,3] ,"red","grey70") )
  
  text( cl_group[cl_group[,3],1] , 
        -log10(cl_group[cl_group[,3],2]) , 
        labels = rownames(cl_group)[cl_group[,3]] , 
        pos=ifelse(cl_group[cl_group[,3],1]>0,3,3) )
  
  abline(v=c(-0.015,0,0.015),h=50,lty=20,col="grey50")
}
dev.off()

```


# Differential cell abundance (VEH controls only, NOT treated with GW )

```{r}
library(glmnet)
library(glmnetUtils)

DATA$orig.ident <- factor(DATA$orig.ident)
temp <- DATA@assays$SCDC_prop_annot@data[,!grepl("GW",DATA$orig.ident)]
group <- factor(gsub(".*[_]","",DATA$orig.ident[!grepl("GW",DATA$orig.ident)]))

plan(multiprocess, workers=future::availableCores()-1  )
temp2 <- data.frame(group)
cl <- future_lapply(rownames(temp),temp=temp,temp2=temp2, function(x,temp,temp2) {
  df <- cbind(temp2,y=temp[x,])
  m <- summary( glm( y ~ group ,data = df) )
  return( m$coefficients )
} , future.seed = 1 )
names(cl) <- rownames(temp)
cl


pdf("differential_abundance_plot_controls_annot.pdf",width = 4*3,height = 4,useDingbats = F)
mypar(1,3)
for( i in rownames(cl[[1]])[-1]){
  
  cl_group <- data.frame(t(sapply(cl,i=i,function(x,i){x[i, c("Estimate","Pr(>|t|)") ]})))
  cl_group <- cbind(cl_group , sig = ( abs(cl_group[,1]) > 0.015 ) & ( cl_group[,2] < 1e-50 ) )
  cl_group[,2][cl_group[,2]==0] <- min(cl_group[,2][cl_group[,2]!=0])

  plot( cl_group[,1] , -log10(cl_group[,2]) , ylim=c(0,400),
        xlim=c(-0.4,0.4), pch=16, main=i, 
        ylab="-log10(pvalue)",xlab="Estimate",
        col= ifelse(cl_group[,3] ,"red","grey70") )
  
  text( cl_group[cl_group[,3],1] , 
        -log10(cl_group[cl_group[,3],2]) , 
        labels = rownames(cl_group)[cl_group[,3]] , 
        pos=ifelse(cl_group[cl_group[,3],1]>0,3,3) )
  
  abline(v=c(-0.015,0,0.015),h=50,lty=20,col="grey50")
}
dev.off()

```




# GSA enrichment


```{r}
modules <- as.data.frame(read.csv2("modulepathways.csv"))
rownames(modules) <- paste0(modules$Module ,"_",modules$Term)

mm <- apply( modules , 1 , tmp = DATA@assays$RNA@data, function(x,tmp){
  comm <- rownames(tmp)[ rownames(tmp) %in% unlist(strsplit(x["Genes"],";")) ]
  temp <- tmp[comm,]
  temp <- temp[rowSums(temp>0)>5,]
  temp <- temp / rowSums(temp)
  temp <- colMeans(temp)
  return(temp)
} )
t(mm)[1,]

colnames(mm) <- rownames(modules)

DATA@assays[['modules_GSA']] <- CreateAssayObject(data = t(mm))
```




# Differential module score

```{r}
library(glmnet)
library(glmnetUtils)

DATA$orig.ident <- factor(DATA$orig.ident)
temp <- DATA@assays$modules_GSA@data[,grepl("d22|d43",DATA$orig.ident)]
group <- factor(gsub(".*[_]","",DATA$orig.ident[grepl("d22|d43",DATA$orig.ident)]))
treat <- factor(gsub("d22|d43","chow",DATA$orig.ident[grepl("d22|d43",DATA$orig.ident)]))


plan(multiprocess, workers=future::availableCores()-1  )

temp2 <- data.frame(group,treat)
cl <- future_lapply(rownames(temp),temp=temp,temp2=temp2, function(x,temp,temp2) {
  df <- cbind(temp2,y=temp[x,])
  m <- summary( glm( y ~ group*treat ,data = df) )
  return( m$coefficients )
} , future.seed = 1 )
names(cl) <- rownames(temp)
cl


pdf("differential_module_score_plot_annot.pdf",width = 4*3,height = 4,useDingbats = F)
mypar(1,3)
for( i in rownames(cl[[1]])[-1]){
  
  cl_group <- data.frame(t(sapply(cl,i=i,function(x,i){x[i, c("Estimate","Pr(>|t|)") ]})))
  cl_group <- cbind(cl_group , sig = ( abs(cl_group[,1]) > 1e-5 ) & ( cl_group[,2] < 1e-2 ) )
  cl_group[,2][cl_group[,2]==0] <- min(cl_group[,2][cl_group[,2]!=0])

  plot( cl_group[,1] , -log10(cl_group[,2]) , ylim=c(0,400),
        xlim=c(-1e-4 ,1e-4 ), pch=16, main=i, 
        ylab="-log10(pvalue)",xlab="Estimate",
        col= ifelse(cl_group[,3] ,"red","grey70") )
  
  text( cl_group[cl_group[,3],1] , 
        -log10(cl_group[cl_group[,3],2]) , 
        labels = gsub( "[-] .*", "",rownames(cl_group)[cl_group[,3]]) , 
        pos=ifelse(cl_group[cl_group[,3],1]>0,3,3) )
  
  abline(v=c(-1e-5,0,1e-5),h=2,lty=2,col="grey50")
}
dev.off()

```


# Differetial module score (VEH controls only, NOT treated with GW )

```{r}
library(glmnet)
library(glmnetUtils)

DATA$orig.ident <- factor(DATA$orig.ident)
temp <- DATA@assays$modules_GSA@data[,!grepl("GW",DATA$orig.ident)]
group <- factor(gsub(".*[_]","",DATA$orig.ident[!grepl("GW",DATA$orig.ident)]))

plan(multiprocess, workers=future::availableCores()-1  )
temp2 <- data.frame(group)
cl <- future_lapply(rownames(temp),temp=temp,temp2=temp2, function(x,temp,temp2) {
  df <- cbind(temp2,y=temp[x,])
  m <- summary( glm( y ~ group ,data = df) )
  return( m$coefficients )
} , future.seed = 1 )
names(cl) <- rownames(temp)


pdf("differential_module_score_plot_control_annot.pdf",width = 4*3,height = 4,useDingbats = F)
mypar(1,3)
for( i in rownames(cl[[1]])[-1]){
  
  cl_group <- data.frame(t(sapply(cl,i=i,function(x,i){x[i, c("Estimate","Pr(>|t|)") ]})))
  cl_group <- cbind(cl_group , sig = ( abs(cl_group[,1]) > 1e-5 ) & ( cl_group[,2] < 1e-2 ) )
  cl_group[,2][cl_group[,2]==0] <- min(cl_group[,2][cl_group[,2]!=0])

  plot( cl_group[,1] , -log10(cl_group[,2]) , ylim=c(0,400),
        xlim=c(-1e-4 ,1e-4 ), pch=16, main=i, 
        ylab="-log10(pvalue)",xlab="Estimate",
        col= ifelse(cl_group[,3] ,"red","grey70") )
  
  text( cl_group[cl_group[,3],1] , 
        -log10(cl_group[cl_group[,3],2]) , 
        labels = gsub( "[-] .*", "",rownames(cl_group)[cl_group[,3]]) , 
        pos=ifelse(cl_group[cl_group[,3],1]>0,3,3) )
  
  abline(v=c(-1e-5,0,1e-5),h=2,lty=2,col="grey50")
}
dev.off()

```






