---
title: "`r params$project_title`"
editor_options:
  chunk_output_type: console
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2:
    code_folding: hide
    toc: yes
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
params:
  dataset: 1
---

```{r Setup, echo = FALSE}
knitr::opts_chunk$set(fig.width  = 10,
                      results    = "hold",
                      message    = FALSE,
                      warning    = FALSE)


```

```{r setup, include=FALSE}

library(Seurat)
library(Matrix)
#remotes::install_github("chris-mcginnis-ucsf/DoubletFinder", upgrade = F)
library(DoubletFinder)
library(ggplot2)


PATH = "/Volumes/jenfra/Private/Projects/Das/sc_irradiation"


# Custom functions

# Basic function to convert human to mouse gene names


convertHumanGeneList <- function(x){
  require("biomaRt")
  human = useMart("ensembl", dataset = "hsapiens_gene_ensembl")
  mouse = useMart("ensembl", dataset = "mmusculus_gene_ensembl")
  
  genesV2 = getLDS(attributes = c("hgnc_symbol"), filters = "hgnc_symbol", values = x , mart = human, attributesL = c("mgi_symbol"), martL = mouse, uniqueRows=T)
  
  humanx <- unique(genesV2[, 2])
  
  # Print the first 6 genes found to the screen
  #print(head(humanx))
  return(humanx)
}

## This is run early due to frequent connection problems with BioMart

# g2mFeat = convertHumanGeneList(cc.genes$g2m.genes)
g2mFeat = c("Smc4","Dlgap5","Ctcf","Cdc25c","Gtse1","Kif20b","Ttk","Tacc3","Top2a","Tpx2","Cdca3","Ncapd2",
"Cks2","Anp32e","G2e3","Lbr","Cdca2","Cks1brt","Ckap2","Hmgb2","Kif11","Nek2","Cenpe","Hjurp",
"Ect2","Aurkb","Cks1b","Kif23","Nuf2","Hmmr","Cdca8","Psrc1","Anln","Cdc20","Birc5","Ndc80",
"Rangap1","Ckap5","Kif2c","Cenpf","Nusap1","Cenpa","Aurka","Ube2c","Ckap2l","Mki67","Tubb4b","Bub1",
"Ccnb2")
#sFeat = convertHumanGeneList(cc.genes$s.genes)
sFeat = c("Msh2","Exo1","Mcm4","Rrm2","Mcm2","Chaf1b","Gmnn","Cdc45","Slbp","Ubr7","Cdc6",
"Rad51ap1","Rpa2","Hells","Fen1","Gins2","Uhrf1","Mcm6","Ung","Dscc1","Usp1","Clspn",
"Cdca7","Pola1","Nasp","Dtl","Mcm5","Wdr76","Prim1","Casp8ap2","Tipin","Blm","Rrm1",
"Brip1","Rad51","Tyms","E2f8","Rfc2","Ccne2","Pcna")

```


# Load data

```{r loading, echo=FALSE}


if(!file.exists("../analysis/original_data.rds")){
  dirs = list.files(PATH, recursive = FALSE)
  datalist = lapply(dirs[grep("_d0",dirs)], function(d){
    CreateSeuratObject(Read10X_h5(filename = paste0(PATH,"/",d,"/outs/filtered_feature_bc_matrix.h5")),
                      d)
  })
  names(datalist) = dirs[grep("_d0",dirs)]
  
  alldata = merge(x = datalist[[1]], y = datalist[-1], add.cell.ids = names(datalist))
  
  unique(alldata$orig.ident)
  saveRDS(alldata, "../analysis/original_data.rds")
  
}else{
  alldata = readRDS("../analysis/original_data.rds")
}



```



# Plot QC vars

Number of features and counts and percentage of mitochondrial and ribosomal RNA are plotted below. Red lines mark limits for filtering.

```{r qc, fig.height=10}



alldata <- PercentageFeatureSet(alldata, "^mt-", col.name = "percent_mito")

alldata <- PercentageFeatureSet(alldata, "^Rp[sl]", col.name = "percent_ribo")

alldata <- PercentageFeatureSet(alldata, "^Hb.-", col.name = "percent_hb")


mito_thresh = 10
ribo_thresh = 5
feat_thresh = 400
hb_thresh = 25


feats <- c("nFeature_RNA", "nCount_RNA", "percent_mito", "percent_ribo", "percent_hb")
glist = VlnPlot(alldata, group.by = "orig.ident", features = feats,  combine = FALSE, pt.size = 0)
names(glist) = feats
glist = lapply(glist, function(x){x + NoLegend() + 
    geom_point(position = position_jitter(), alpha = 0.1, size = 0.1)})


cowplot::plot_grid(align = "v",plotlist = list(glist[["nFeature_RNA"]] + geom_hline(yintercept = feat_thresh, color = "red"),
glist[["nCount_RNA"]],
glist[["percent_mito"]] + geom_hline(yintercept = mito_thresh, color = "red") ,
glist[["percent_ribo"]] + geom_hline(yintercept = ribo_thresh, color = "red")),
glist[["percent_hb"]] + geom_hline(yintercept = hb_thresh, color = "red"))


FeatureScatter(alldata, "nCount_RNA", "nFeature_RNA", group.by = "orig.ident", pt.size = 0.5,
                shuffle = TRUE) + geom_hline(yintercept = feat_thresh)

FeatureScatter(alldata, "percent_ribo", "percent_mito", group.by = "orig.ident", pt.size = 0.5,
                shuffle = TRUE) + geom_hline(yintercept = mito_thresh) + 
  geom_hline(yintercept = mito_thresh) +
  geom_vline(xintercept = ribo_thresh)

FeatureScatter(subset(alldata, idents = c("GW_EPI_d0", "STD_EPI_d0")),
               "percent_ribo", "percent_mito", group.by = "orig.ident",
               pt.size = 0.5, shuffle = TRUE, raster = TRUE) + 
  geom_hline(yintercept = mito_thresh) +
  geom_vline(xintercept = ribo_thresh)

```

# Filter cells and genes

The same QC as above are plotted below, only including cells and genes that pass filtering limits (gene limit: > 3 copies expressed). The top expressed genes are also shown. Mitochondrial and ribosomal genes are filtered out of the data. 

```{r filtering, fig.height = 10}

selected_c <- WhichCells(alldata, expression = nFeature_RNA > feat_thresh)
selected_f <- rownames(alldata)[Matrix::rowSums(alldata) > 3]

data.filt <- subset(alldata, features = selected_f, cells = selected_c)

selected_mito <- WhichCells(data.filt, expression = percent_mito < 10)
selected_ribo <- WhichCells(data.filt, expression = percent_ribo > 5)
selected_hb <- WhichCells(data.filt, expression = percent_hb < 25)

# and subset the object to only keep those cells
data.filt <- subset(data.filt, cells = intersect(selected_hb,intersect(selected_mito, selected_ribo)))

feats <- c("nFeature_RNA", "nCount_RNA", "percent_mito", "percent_ribo")
VlnPlot(data.filt, group.by = "orig.ident", features = feats, pt.size = 0.1, ncol = 2) + 
    NoLegend()
VlnPlot(data.filt, group.by = "orig.ident", features = feats, pt.size = 0, ncol = 2) + 
    NoLegend()



par(mar = c(4, 8, 2, 1))
C <- data.filt@assays$RNA@counts
C <- Matrix::t(Matrix::t(C)/Matrix::colSums(C)) * 100

most_expressed <- order(apply(C, 1, median), decreasing = T)[20:1]

# potentials = list()
# for(i in seq(1,dim(C)[2],20000)){
#   potentials[[length(potentials)+1]] = rownames(C)[order(apply(C[,i:min(i+19999, dim(C)[2])], 1, median), decreasing = T)[1000:1]]
# }
# length(unique(unlist(potentials)))
# 
# 
# most_expressed <- order(apply(C[unique(unlist(potentials)),], 1, median), decreasing = T)[20:1]


boxplot(as.matrix(t(C[most_expressed, ])), cex = 0.1, las = 1, xlab = "% total count per cell",
    col = (scales::hue_pal())(20)[20:1], horizontal = TRUE)

# Filter Malat1, Gm42418
data.filt <- data.filt[!grepl("Malat1", rownames(data.filt)), ]
data.filt <- data.filt[!grepl("Gm42418", rownames(data.filt)), ]

# Filter Mitocondrial
data.filt <- data.filt[!grepl("^mt-", rownames(data.filt)), ]

# Filter Ribossomal gene (optional if that is a problem on your data) data.filt
data.filt <- data.filt[ ! grepl('^Rp[sl]', rownames(data.filt)), ]

# Filter Hemoglobin gene (optional if that is a problem on your data)
data.filt <- data.filt[!grepl("^Hb.-", rownames(data.filt)), ]

# saveRDS(data.filt, "../analysis/data_pre_cellcycle.rds")

``` 

# Analyze cell cycle

The S-phase and G2M-phase scores are calculated based on the cell cycle gene lists defined in the Seurat package (translated to mouse genes using biomaRt).

```{r cellcycle}
# Before running CellCycleScoring the data need to be normalized and
# logtransformed.

data.filt = NormalizeData(data.filt)

data.filt <- CellCycleScoring(object = data.filt, g2m.features = g2mFeat, 
    s.features = sFeat)

VlnPlot(data.filt, features = c("S.Score", "G2M.Score"), group.by = "orig.ident", 
    ncol = 3, pt.size = 0.1)

```


# Find Doublets

Doublets are identified and excluded using the DoubletFinder package.

```{r doublets}

if(file.exists(( "../analysis/filtered_data_presweep.rds"))){
  data.filt = readRDS( "../analysis/filtered_data_presweep.rds")
}else{
  data.filt = FindVariableFeatures(data.filt, verbose = F)
  data.filt = ScaleData(data.filt, vars.to.regress = c("nFeature_RNA", "percent_mito"), 
      verbose = F)
  data.filt = RunPCA(data.filt, verbose = F, npcs = 20)
  data.filt = RunUMAP(data.filt, dims = 1:10, verbose = F)
  
  saveRDS(data.filt, "../analysis/filtered_data_presweep.rds")
}




if(file.exists("../analysis/sweeplist.RData")){
  load("../analysis/sweeplist.RData")
}else{
  sweeplist=list()
  for(i in unique(data.filt$orig.ident)){
    data.filt.subset=subset(data.filt, idents = i)
    sweep.res <- paramSweep_v3(data.filt.subset)#, num.cores = 3)
    sweep.stats <- summarizeSweep(sweep.res, GT = FALSE)
    bcmvn <- find.pK(sweep.stats)
    
    sweeplist[[i]] = bcmvn
    
  }
  
  save(sweeplist, file = "../analysis/sweeplist.RData")
}


if(file.exists("../analysis/singletList.RData")){
  load("../analysis/singletList.RData")
}else{
  singletList = list()
  
  
  for(i in unique(data.filt$orig.ident)){
    data.filt.subset=subset(data.filt, idents = i)
    nExp <- round(ncol(data.filt.subset) * 0.08)  # expect 8% doublets
    mypK = as.numeric(as.character(sweeplist[[i]]$pK[which.max(sweeplist[[i]]$BCmetric)]))
    data.filt.subset <- doubletFinder_v3(data.filt.subset, pN = 0.25, pK = mypK, nExp = nExp, PCs = 1:10)
    DF.name = colnames(data.filt.subset@meta.data)[grepl("DF.classification", colnames(data.filt.subset@meta.data))]
    singletList[[i]] = data.filt.subset@meta.data[, DF.name]
    names(singletList[[i]]) = colnames(data.filt.subset)
  }
  
  save(singletList, file = "../analysis/singletList.RData")
}


# sweep.res <- paramSweep_v3(data.filt)#, num.cores = 3)
# sweep.stats <- summarizeSweep(sweep.res, GT = FALSE)
# bcmvn <- find.pK(sweep.stats)
# par(mar = c(4, 4, 2, 2))
# barplot(bcmvn$BCmetric, names.arg =bcmvn$pK, las=2)





#data.filt@meta.data$DF = unlist(singletList)[match(colnames(data.filt), gsub(".*\\.","",names(unlist(singletList))))]
data.filt = AddMetaData(data.filt, unlist(singletList)[match(colnames(data.filt), gsub(".*\\.","",names(unlist(singletList))))], "DF")

cowplot::plot_grid(ncol = 2, DimPlot(data.filt, group.by = "orig.ident") + NoAxes(), 
    DimPlot(data.filt, group.by = "DF") + NoAxes())


DimPlot(data.filt, group.by = "DF", split.by = "orig.ident", ncol = 4) + NoAxes()




VlnPlot(data.filt, features = "nFeature_RNA", group.by = "DF",pt.size = 0.1)
VlnPlot(data.filt, features = "nFeature_RNA", split.by = "DF", group.by = "orig.ident", pt.size = 0.1)


data.filt = data.filt[, data.filt@meta.data[, "DF"] == "Singlet"]



cellfreq = as.data.frame(table(data.filt$orig.ident), stringsAsFactors = FALSE)
cellfreq$Day = sapply(cellfreq$Var1, function(x){strsplit(x, "_")[[1]][3]})
cellfreq$Type = sapply(cellfreq$Var1, function(x){strsplit(x, "_")[[1]][2]})
cellfreq$Treatment = sapply(cellfreq$Var1, function(x){strsplit(x, "_")[[1]][1]})
#png("/Volumes/tank/Private/Meetings/Lab meetings/MyLabMeetings/20210621/FilteredCellNumbers.png", res = 240, width = 1200, height = 800)
ggplot(cellfreq, aes(x = paste(Day, Treatment), y = Freq, color = Type)) + geom_point() + 
  theme(panel.background = element_blank(), legend.key = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1)) + 
  labs(x = "Day and treatment", y = "Number of cells") +
  ylim(0,NA)
#dev.off()



```

Final cell counts:
`r knitr::kable(table(data.filt$orig.ident))`


```{r final}

saveRDS(data.filt, file = "../analysis/filtered_data.rds")


sessionInfo()

```




