---
title: "`r params$project_title`"
editor_options:
  chunk_output_type: console
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: yes
  bookdown::html_document2:
    code_folding: hide
    toc: yes
    toc_float:
      collapsed: no
params:
  dataset: 1
---

```{r Setup, echo = FALSE}
knitr::opts_chunk$set(fig.width  = 7,
                      results    = "hold",
                      message    = FALSE,
                      warning    = FALSE)


```

```{r setup}

suppressPackageStartupMessages({
    library(Seurat)
    library(venn)
    library(dplyr)
    library(cowplot)
    library(ggplot2)
    library(pheatmap)
    library(enrichR)
    library(rafalib)
    library(cowplot)
  library(gplots) 
  library(ggrepel)
})


plottheme = theme(legend.title = element_text(size = 8),
                  legend.text = element_text(size = 8),
                  axis.title = element_text(size = 8),
                  axis.text = element_text(size = 8),
                  plot.title = element_text(size = 8))


# library(future)
# plan("multiprocess", workers = 4)
# options(future.globals.maxSize = 24000 * 1024^2)

alldata <- readRDS("../analysis/filtered_EPI_int_clus.rds")

# Set desired clustering resolution
sel.lev = 1

# Set the identity as louvain with resolution selected above
sel.clust = paste0("CCA_snn_res.",sel.lev)

alldata$orig.ident = factor(alldata$orig.ident, levels = c("STD_EPI_d0","GW_EPI_d0"))
alldata$Treatment = factor(alldata$Treatment, levels = c("STD","GW"))

alldata@active.assay = "RNA"

alldata <- SetIdent(alldata, value = sel.clust)
table(alldata@active.ident)
alldata$ident_clust = paste0(alldata@active.ident,"_",alldata$orig.ident)


ggplotColours <- function(n = 6, h = c(0, 360) + 15){
  if ((diff(h) %% 360) < 1) h[2] <- h[2] - 360/n
  hcl(h = (seq(h[1], h[2], length = n)), c = 100, l = 65)
}

cluster_colors = ggplotColours(length(unique(alldata@active.ident))+1)[-1]

```

# Used clustering



```{r clusteringdim}

plot_grid(ncol = 2, DimPlot(alldata, label = T, shuffle = TRUE, 
                                        raster = TRUE) + 
            NoAxes() + NoLegend() +
            scale_color_manual(values = cluster_colors),
          DimPlot(alldata, group.by = "orig.ident", shuffle = TRUE, 
                                        raster = TRUE) +
            scale_color_manual(values = rev(ggplotColours(2))) + 
            NoAxes() + theme(legend.position = "bottom"))

```

```{r, fig.height =4}




frequencies = as.data.frame(table(data.frame(cluster=alldata@meta.data[,sel.clust],
                 sample = alldata$orig.ident)))
frequencies$Proportion=NA
for(i in 1:nrow(frequencies)){
  frequencies$Proportion[i] = frequencies$Freq[i]/sum(frequencies$Freq[frequencies$sample==frequencies$sample[i]])
}

plot_grid(ggplot(alldata@meta.data, aes_string(fill= sel.clust, x = "Treatment")) +
            plottheme + 
            geom_bar(position = "fill", color = "black") +
            theme(panel.background = element_blank(),
                  axis.ticks = element_blank(), 
                  axis.title.y = element_blank()),
          ggplot(frequencies, aes(fill=sample,
                                  y = Proportion, 
                                  x = cluster)) + plottheme + 
            geom_bar(stat = "identity",position="dodge") +
            theme(panel.background = element_blank(),
                  axis.ticks.y = element_blank()),
          rel_widths = c(2,3))


```


```{r, fig.height = 3}

DotPlot(alldata, features = c("Ptprc","Epcam"), 
    assay = "RNA") + coord_flip() + plottheme

```


# Cluster marker genes

Top 25 (by p-value) genes with logFC>0.2 for each cluster, sorted by logFC:


```{r,  fig.height = 10}




if(file.exists(paste0("../analysis/clustermarkers_EPI_CCA_",sel.lev,".rds"))){
  markers_genes = readRDS(paste0("../analysis/clustermarkers_EPI_CCA_",sel.lev,".rds"))
}else{
  markers_genes <- FindAllMarkers(alldata, logfc.threshold = 0.2, test.use = "wilcox", 
  min.pct = 0.1, min.diff.pct = 0.2, only.pos = TRUE, max.cells.per.ident = 300, 
  assay = "RNA", random.seed = 42)
  saveRDS(markers_genes, paste0("../analysis/clustermarkers_EPI_CCA_",sel.lev,".rds"))
}


top25 <- markers_genes %>% group_by(cluster) %>% top_n(-25, p_val)
top25 <- top25 %>% group_by(cluster) %>% top_n(25, avg_log2FC)


mypar(2, 5, mar = c(4, 6, 3, 1))
for (i in unique(top25$cluster)) {
    barplot(sort(setNames(top25$avg_log2FC, top25$gene)[top25$cluster == i], F), horiz = T, 
        las = 1, main = paste0(i, " vs. rest"), border = "white", yaxs = "i")
    abline(v = c(0, 0.25), lty = c(1, 2))
}



```


Top 5 (by p-value) genes with logFC>0.2 for each cluster:

```{r, fig.height=8}

top5 <- markers_genes %>% group_by(cluster) %>% top_n(-5, p_val_adj)
top5 <- top5 %>% group_by(cluster) %>% top_n(5, avg_log2FC)

alldata <- ScaleData(alldata, features = as.character(unique(top5$gene)), assay = "RNA")
DoHeatmap(alldata, features = as.character(unique(top5$gene)), group.by = sel.clust, 
    assay = "RNA") + plottheme

#pdf(paste0("../figures/clustermarkers_EPI_",sel.lev,".pdf"), width = 10, height = 10)
#DoHeatmap(alldata, features =  as.character(unique(top5$gene)), assay ="RNA",label = FALSE) + labs(color = "Cluster")
#dev.off()


DotPlot(alldata, features = rev(as.character(unique(top5$gene))), 
    assay = "RNA") + coord_flip() + plottheme


# topAll = markers_genes %>% group_by(cluster)
# DoHeatmap(alldata, features =  as.character(unique(topAll$gene)), assay ="RNA",label = FALSE) + labs(color = "Cluster") 


# DoHeatmap(alldata, features = as.character(unique(top5$gene)), group.by = sel.clust,
#     assay = "RNA", label = FALSE)
# 
# DotPlot(alldata, features = rev(as.character(unique(top5$gene))), group.by = sel.clust, 
#     assay = "RNA") + coord_flip()

```

Top 3 (by p-value) genes with logFC>0.2 for each cluster:

```{r, fig.height=10}

top3 <- top5 %>% group_by(cluster) %>% top_n(-3, p_val)
top3 <- top3 %>% group_by(cluster) %>% top_n(3, avg_log2FC)

VlnPlot(alldata, features = as.character(unique(top3$gene)), ncol = 5, group.by = sel.clust, 
    assay = "RNA", pt.size = -1)

```

```{r, fig.height=12}

plot_grid(plotlist = lapply(FeaturePlot(alldata, 
                                        features = as.character(unique(top3$gene)), 
                                        raster = TRUE,
    slot = "data", combine = FALSE), function(x){x + NoAxes() + plottheme + 
        theme(legend.key.width = unit(5, "pt"))}), ncol = 4)

```


# Annotations

Based on the top DE genes, here are my predctions for the cell types represented by each cluster:

* Cluster 0: Enterocytes
* Cluster 1: Immune cells
* Cluster 2: Stem cells
* Cluster 3: Proliferating stem cells
* Cluster 4: Immune cells
* Cluster 5: Other (low quality enterocytes, enteroendocrine)
* Cluster 6: Enterocytes (Apoc2, Apoa1, Aldh1a1)
* Cluster 7: Goblet cells

# Diet comparisons

## Full samples

```{r, fig.height = 3}

volcanotheme = theme(axis.text = element_text(size = 8),
                     axis.title = element_text(size = 8),
                     panel.grid = element_blank(),
                     panel.background = element_blank(),
                     
                    plot.title = element_text(hjust = 0.5,
                                              size = 10))

if(!file.exists("../analysis/epi_de_all.rds")){
epi_de = FindMarkers(alldata,
            ident.1 = "GW",ident.2 = "STD",
            group.by = "Treatment", logfc.threshold = -Inf)
saveRDS(epi_de, "../analysis/epi_de_all.rds")
}else{
  epi_de = readRDS("../analysis/epi_de_all.rds")
}

epi_de$gene = rownames(epi_de)


vdata = epi_de
g_all = ggplot(vdata, aes(x = avg_log2FC, y = -log10(p_val_adj))) + 
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) + 
  geom_point(size = 1) + 
  geom_text_repel(data = vdata[abs(vdata$avg_log2FC)>1,],
                  aes(label = gene), size = 2.5) + 
  labs(title = "All cells") +
  volcanotheme


epi_de_500 = FindMarkers(alldata,
            ident.1 = "GW",ident.2 = "STD",
            group.by = "Treatment", logfc.threshold = -Inf,
            max.cells.per.ident = 500)

epi_de_500$gene = rownames(epi_de_500)

vdata = epi_de_500
g_500 = ggplot(vdata, aes(x = avg_log2FC, y = -log10(p_val_adj))) + 
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) + 
  geom_point(size = 1) + 
  geom_text_repel(data = vdata[abs(vdata$avg_log2FC)>1,],
                  aes(label = gene), size = 2.5) + 
  labs(title = "500 cells per group") +
  volcanotheme

plot_grid(g_all,
          g_500)

```


## Per cluster

### All cells

For each cluster, the differentially expressed genes between the two treatments are identified. The results are shown below: Dot plots for the top 20 DE genes and volcano plots for all DE genes. Note that the higher significance in the first clusters is primarily due to the sizes of the clusters.

```{r de, fig.height = 4, fig.width = 2.5}


if(!file.exists(paste0("../analysis/DE_GWvsSTD_EPI_",sel.lev,".rds"))){
DGE_list_Treatment = list()

  for(i in sort(unique(alldata@active.ident))){
  message(paste("Cluster",i))
  # select all cells in cluster i
  cell_selection <- subset(alldata, cells = colnames(alldata)[alldata@active.ident == 
      i])
  cell_selection <- SetIdent(cell_selection, value = "Treatment")
  # Compute differential expression
  DGE_cell_selection <- FindMarkers(ident.1 = "GW", ident.2 = "STD",
                                    cell_selection, logfc.threshold = 0.2,
                                    test.use = "wilcox", 
      min.pct = 0.1, only.pos = FALSE,
      assay = "RNA")
  DGE_list_Treatment[[paste("Cluster",i)]] = DGE_cell_selection[DGE_cell_selection$p_val_adj<0.001,]
  DGE_list_Treatment[[paste("Cluster",i)]]$gene = rownames(DGE_list_Treatment[[paste("Cluster",i)]])
  saveRDS(DGE_list_Treatment, paste0("../analysis/DE_GWvsSTD_EPI_",sel.lev,".rds"))
}}else{
  DGE_list_Treatment = readRDS(paste0("../analysis/DE_GWvsSTD_EPI_",sel.lev,".rds"))
}

```


```{r}

num.cells = as.matrix(table(alldata@active.ident,
                  alldata$Treatment))
rownames(num.cells) = paste0("Cluster ", rownames(num.cells))

reglist = lapply(names(DGE_list_Treatment), function(i){
  message(i)
  x = DGE_list_Treatment[[i]]
  c(up = sum(x$avg_log2FC > 0),
    down = sum(x$avg_log2FC < 0),
    total = nrow(x),
    min.cells = min(num.cells[i,]))})
names(reglist) = names(DGE_list_Treatment)
regdf = as.data.frame(reglist)
colnames(regdf) = gsub("Cluster.","", colnames(regdf))

```

Number of DE genes per cluster

`r knitr::kable(regdf)`

Below are the numbers of DE genes per cluster according to size of cluster. 

```{r,fig.height = 4, fig.width = 4}


tregdf = as.data.frame(t(regdf))
tregdf$Cluster = rownames(tregdf)
ggplot(tregdf, aes(x = min.cells, y = total)) + 
  geom_text(aes(label = Cluster ))

```

```{r dotplots, fig.height = 4, fig.width = 2.5}
for(i in sort(unique(alldata@active.ident))){
  top20_cell_selection <- DGE_list_Treatment[[paste("Cluster",i)]]
  n = min(nrow(top20_cell_selection), 20)
  top20_cell_selection <- top20_cell_selection[order(top20_cell_selection$p_val)[1:n],]
  top20_cell_selection <- top20_cell_selection[order(top20_cell_selection$avg_log2FC)[1:n],]
  # print(VlnPlot(subset(alldata, idents = i),unique(top20_cell_selection$gene), 
  #         group.by = "orig.ident", assay = "RNA") + theme(axis.text = element_text(size = 1)))
  if(nrow(top20_cell_selection)>0){
    d = DotPlot(subset(alldata, idents = i), scale = FALSE, 
                features = rev(as.character(unique(top20_cell_selection$gene))), 
      assay = "RNA", group.by = "Treatment") + coord_flip() + plottheme +
        labs(title = paste("Cluster (",i,paste(top3$gene[top3$cluster==i],collapse = ", "),
                           ") (",nrow(DGE_list_Treatment[[paste("Cluster",i)]])," genes)"))
    d$guides$size$title = "Percent"
    d$guides$colour$title = "Average"
    print(d)
  }
}

```

```{r volcanos, fig.height = 4, fig.width = 7}


for(i in sort(unique(alldata@active.ident))){
  res = DGE_list_Treatment[[paste("Cluster",i)]]
  
  #res$avg_log2FC[res$cluster=="STD"] = 0-res$avg_log2FC[res$cluster=="STD"]
  
  if(nrow(res)>0){
      print(ggplot(res, aes(x = avg_log2FC, y = -log10(p_val_adj))) +
              geom_point(color = "#AAAAAA") + 
              geom_hline(yintercept = 0) +
              geom_vline(xintercept = 0) + 
              geom_text_repel(aes(label = gene), size = 2.5) +
              plottheme +
              theme(panel.background = element_blank(),
                    panel.grid = element_blank()) + 
        labs(title = paste("Cluster",i,paste(top3$gene[top3$cluster==i],collapse = ", "))))
  
  }
}



```



### Max 100 cells per group

To enable comparison between DE in different clusters, the DE analysis is redone using a smaller number of cells (100 per group). The results are shown in the same format as above.

```{r de_100, fig.height = 4, fig.width = 2.5}


if(!file.exists(paste0("../analysis/DE_GWvsSTD_EPI_",sel.lev,"_balanced.rds"))){
DGE_list_Treatment = list()

  for(i in sort(unique(alldata@active.ident))){
  message(paste("Cluster",i))
  # select all cells in cluster i
  cell_selection <- subset(alldata, cells = colnames(alldata)[alldata@active.ident == 
      i])
  cell_selection <- SetIdent(cell_selection, value = "Treatment")
  # Compute differential expression
  DGE_cell_selection <- FindMarkers(ident.1 = "GW", ident.2 = "STD",
                                    cell_selection, logfc.threshold = 0.2,
                                    test.use = "wilcox", max.cells.per.ident = 100,
      min.pct = 0.1, only.pos = FALSE,
      assay = "RNA")
  DGE_list_Treatment[[paste("Cluster",i)]] = DGE_cell_selection[DGE_cell_selection$p_val_adj<0.001,]
  DGE_list_Treatment[[paste("Cluster",i)]]$gene = rownames(DGE_list_Treatment[[paste("Cluster",i)]])
  saveRDS(DGE_list_Treatment, paste0("../analysis/DE_GWvsSTD_EPI_",sel.lev,"_balanced.rds"))
}}else{
  DGE_list_Treatment = readRDS(paste0("../analysis/DE_GWvsSTD_EPI_",sel.lev,"_balanced.rds"))
}
```


```{r}

num.cells = as.matrix(table(alldata@active.ident,
                  alldata$Treatment))
num.cells[num.cells>100] = 100
rownames(num.cells) = paste0("Cluster ", rownames(num.cells))

reglist = lapply(names(DGE_list_Treatment), function(i){
  message(i)
  x = DGE_list_Treatment[[i]]
  c(up = sum(x$avg_log2FC > 0),
    down = sum(x$avg_log2FC < 0),
    total = nrow(x),
    min.cells = min(num.cells[i,]))})
names(reglist) = names(DGE_list_Treatment)
regdf = as.data.frame(reglist)
colnames(regdf) = gsub("Cluster.","", colnames(regdf))

```

Number of DE genes per cluster

`r knitr::kable(regdf)`


```{r,fig.height = 4, fig.width = 4}


tregdf = as.data.frame(t(regdf))
tregdf$Cluster = rownames(tregdf)
ggplot(tregdf, aes(x = as.numeric(Cluster), y = total)) + 
  geom_bar(stat = "identity")

```

```{r dotplots_100, fig.height = 4, fig.width = 2.5}

for(i in sort(unique(alldata@active.ident))){
  top20_cell_selection <- DGE_list_Treatment[[paste("Cluster",i)]]
  n = min(nrow(top20_cell_selection), 20)
  top20_cell_selection <- top20_cell_selection[order(top20_cell_selection$p_val)[1:n],]
  top20_cell_selection <- top20_cell_selection[order(top20_cell_selection$avg_log2FC)[1:n],]
  # print(VlnPlot(subset(alldata, idents = i),unique(top20_cell_selection$gene), 
  #         group.by = "orig.ident", assay = "RNA") + theme(axis.text = element_text(size = 1)))
  if(nrow(top20_cell_selection)>0){
    d = DotPlot(subset(alldata, idents = i), scale = FALSE, 
                features = rev(as.character(unique(top20_cell_selection$gene))), 
      assay = "RNA", group.by = "Treatment") + coord_flip() + plottheme +
        labs(title = paste("Cluster (",i,paste(top3$gene[top3$cluster==i],collapse = ", "),
                           ") (",nrow(DGE_list_Treatment[[paste("Cluster",i)]])," genes)"))
    d$guides$size$title = "Percent"
    d$guides$colour$title = "Average"
    print(d)
  }
}

```

```{r volcanos_100, fig.height = 4, fig.width = 7}


for(i in sort(unique(alldata@active.ident))){
  res = DGE_list_Treatment[[paste("Cluster",i)]]
  
  #res$avg_log2FC[res$cluster=="STD"] = 0-res$avg_log2FC[res$cluster=="STD"]
  
  if(nrow(res)>0){
      print(ggplot(res, aes(x = avg_log2FC, y = -log10(p_val_adj))) +
              geom_point(color = "#AAAAAA") + 
              geom_hline(yintercept = 0) +
              geom_vline(xintercept = 0) + 
              geom_text_repel(aes(label = gene), size = 2.5) +
              plottheme +
              theme(panel.background = element_blank(),
                    panel.grid = element_blank()) + 
        labs(title = paste("Cluster",i,paste(top3$gene[top3$cluster==i],collapse = ", "))))
  
  }
}



```
